---
title: "Label-free quantification with FlashLFQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Label-free quantification with FlashLFQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge running FlashLFQ on
# mzLib's own test data, so the page builds anywhere. With a bridge installed, the same calls
# quantify your runs.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

[FlashLFQ](https://github.com/smith-chem-wisc/FlashLFQ) is the label-free quantification engine
MetaMorpheus uses. Give it a search result and the mzML runs it came from, and it measures how
much of each peptide and protein is in each run. The whole pipeline is mzLib's: its readers read
the result file, its converter makes FlashLFQ identifications, and `FlashLfqEngine` quantifies
them. MetaMorpheus is not involved.

| question | function | mzLib |
|---|---|---|
| how much of each peptide and protein is in each run? | `flashlfq_quantify()` | `FlashLfqEngine.Run` |
| fill in peptides missing from a run | `flashlfq_quantify(match_between_runs = TRUE)` | FlashLFQ's match-between-runs |
| group runs into conditions and replicates | `flashlfq_quantify(spectra = <data.frame>)` | `SpectraFileInfo` |
| how many peaks did match-between-runs transfer? | `flashlfq_mbr_peaks()`, `flashlfq_mbr_peak_count()` | the peak table |
| re-roll proteins from a peptide table, without the mzML | `flashlfq_median_polish()` | `FlashLfqResults.CalculateProteinResultsMedianPolish` |

The data are real: two K562 runs from mzLib's FlashLFQ test data and their MetaMorpheus search,
quantified with match-between-runs on.

```{r}
runs <- c("20100614_Velos1_TaGe_SA_K562_3.mzML", "20100614_Velos1_TaGe_SA_K562_4.mzML")
quant <- flashlfq_quantify("AllPSMs.psmtsv", runs, match_between_runs = TRUE)
quant
```

Any quantifiable search result works, not only MetaMorpheus's: an MSFragger `psm.tsv` or a DIA-NN
`report.tsv` too. mzLib converts MSFragger's retention times from seconds to minutes as it reads
them (mzLib #1116), so FlashLFQ searches the right window.

## Four tidy tables

```{r}
quant$spectra_files[, c("file_name", "peak_count", "mbr_peak_count")]
head(quant$peptides, 4)
head(quant$proteins[quant$proteins$intensity > 0 & !is.na(quant$proteins$intensity), ], 4)
head(quant$peaks[, c("file_name", "sequence", "intensity", "detection_type", "retention_time")], 4)
```

`peptides` and `proteins` are long: one row per peptide or protein group per run, with the run
in `file_name`. Intensities are in the instrument's units; `retention_time` in `peaks` is the
apex, in minutes.

Most protein groups read 0 in both runs, and that is a setting, not a failure. With
`use_shared_peptides_for_protein_quant = FALSE`, the default, only a group's own peptides count,
and most groups here have none.

## 0 and `NA` mean different things

A **peptide** intensity of `0` means the peptide was not measured in that run. A **protein**
intensity of `NA` means FlashLFQ could not resolve a number at all. mzLibR keeps the two apart,
so `mean()` over a protein column returns `NA` rather than a confidently wrong number, and
`na.rm = TRUE` is a choice you make visibly.

```{r}
c(not_resolved = sum(is.na(quant$proteins$intensity)),
  not_measured = sum(quant$proteins$intensity == 0, na.rm = TRUE))
```

## Count transfers from the peaks

With `match_between_runs = TRUE`, a peptide identified in one run but missing from another is
still quantified in the second, by transferring the identification across the aligned
retention-time axis. Count transfers from `peaks`, never from the peptide roll-up, which mirrors
FlashLFQ's `QuantifiedPeptides.tsv` and drops most of them:

```{r}
flashlfq_mbr_peak_count(quant)
table(quant$peptides$file_name[quant$peptides$detection_type == "MBR"])
head(flashlfq_mbr_peaks(quant)[, c("file_name", "sequence", "intensity")], 3)
```

None of run 3's transfers appear in the peptide table: they read `"NotDetected"` with intensity
0. A peptide by run matrix built from `peptides` silently drops most of what MBR filled in; build
it from `peaks`.

## Re-quantify proteins under a new design

`flashlfq_median_polish()` reruns only the protein roll-up, from the `QuantifiedPeptides.tsv`
FlashLFQ wrote, without re-reading any spectra. With no design, each run is its own sample, and
the answer is FlashLFQ's own:

```{r}
polished <- flashlfq_median_polish("K562_QuantifiedPeptides.tsv")
polished$samples
key <- function(groups, runs) paste(groups, runs)
same <- match(key(quant$proteins$protein_group, quant$proteins$file_name),
              key(polished$proteins$protein_group, polished$proteins$sample))
identical(quant$proteins$intensity, polished$proteins$intensity[same])
```

Every protein group's intensity in every run is reproduced exactly. Give `design` a data.frame with
`file_name`, `condition` and `biological_replicate` to say which runs are replicates of which
sample, and only the roll-up is redone.

## Before a real run

- Use a MetaMorpheus `.psmtsv` or `.osmtsv`, an MSFragger `psm.tsv` or a DIA-NN report, and mzML
  runs; convert `.raw` first.
- Give `spectra` a `condition` and `biological_replicate` so FlashLFQ knows which runs are
  comparable; match-between-runs needs a balanced design. `sdrf_design_spectra()` builds that
  from an SDRF.
- `max_threads` defaults to 1 here and to every core in pyMzLib and mzLibRust;
  `?flashlfq_quantify` says why.

## What to cite

The methods and resources behind the functions on this page, from their specs:

<!-- BEGIN cite: generated by scripts/render-cite.R from the verb specs; do not edit -->

- [doi:10.1021/acs.jproteome.7b00608](https://doi.org/10.1021/acs.jproteome.7b00608): FlashLFQ: the peak-finding, match-between-runs and protein quantification this verb runs (Millikin et al., J Proteome Res 2018).

<!-- END cite -->

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
