---
title: "Isobaric kits: TMT, TMTpro, iTRAQ and DiLeu channels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Isobaric kits: TMT, TMTpro, iTRAQ and DiLeu channels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, so the page builds
# anywhere. With a bridge installed, the same calls ask mzLib.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

An isobaric experiment - TMT, TMTpro, iTRAQ, DiLeu - is read out of the low-m/z **reporter
ions**: one ion per channel, some only millidaltons apart. Mislabel one channel, or shift its m/z,
and every sample after it is assigned to the wrong condition, with nothing in the result looking
wrong. So mzLibR keeps no table of its own. `isobaric_kits()` returns **mzLib's**, the one
MetaMorpheus quantifies TMT searches with:

| question | function | mzLib |
|---|---|---|
| every kit mzLib knows, with every channel | `isobaric_kits()` | `IsobaricMassTag`, `IsobaricMassTagType` (#1375) |
| one kit, by its name or a MetaMorpheus modification name | `isobaric_kits("TMT18")` | `IsobaricMassTag.TryGetTagType` |
| a channel's reporter-ion m/z and matching window | `records$reporter_ion_mz`, `mz_min`, `mz_max` | `ReporterIonMzs`, `ReporterIonMzRanges` |

## Every kit

```{r}
catalogue <- isobaric_kits()
catalogue
catalogue$kits
```

`records` is one row per (kit, channel). `channel_index` is 1-based within its kit, ascending with
the m/z.

**`TMT16` and `TMT18` are TMTpro.** TMT16 is the lowest sixteen channels of the 18-plex set and
has no modification entry of its own, so its m/z are TMT18's first sixteen. **iTRAQ 8-plex has no
120 channel**: the phenylalanine immonium ion sits at m/z 120.081, so the eighth reagent is 121.

```{r}
catalogue$records$channel_label[catalogue$records$kit == "iTRAQ8"]
```

## One kit: TMTpro 18-plex

Ask for a kit the way MetaMorpheus names it. The match is on the **whole name**,
case-insensitive, optionally followed by `" on <motif>"`: `"TMT18"`, `"tmt10"`, `"TMT6-plex"`
and `"iTRAQ-4plex on K"` all work. A substring never does, so `"TMT10plex"` is refused with a
list of the kits that exist, rather than quietly resolved to whichever kit's name it contains.

```{r}
tmtpro <- isobaric_kits("TMT18")
head(tmtpro$records[, c("channel_index", "channel_label", "reporter_ion_mz", "mz_min", "mz_max")], 4)
```

## What the numbers are

Each `reporter_ion_mz` is a `DI HCD` diagnostic-ion line of the kit's `Multiplex Label`
modification in mzLib's embedded `TMT.txt`, plus one proton: theoretical, at charge 1, never
calibrated or observed. mzLib reads a reporter intensity as the most intense peak between
`mz_min` and `mz_max`, `absolute_tolerance` Da either side:

```{r}
tmtpro$absolute_tolerance
```

mzLib takes that tolerance from the TMTpro 18-plex paper (Li et al., J Proteome Res 2021,
[doi:10.1021/acs.jproteome.1c00168](https://doi.org/10.1021/acs.jproteome.1c00168)), as its
`IsobaricMassTag.AbsoluteToleranceValue` says.

## What to cite

The methods and resources behind the functions on this page, from their specs:

<!-- BEGIN cite: generated by scripts/render-cite.R from the verb specs; do not edit -->

- [doi:10.1021/acs.jproteome.1c00168](https://doi.org/10.1021/acs.jproteome.1c00168): the 0.003 Da reporter-ion matching tolerance mzLib uses (IsobaricMassTag.AbsoluteToleranceValue).

<!-- END cite -->

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
