---
title: "Digesting and fragmenting peptidoforms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Digesting and fragmenting peptidoforms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge fetching UniProt, so
# the page builds offline. With a bridge installed, the same calls fetch the entry live.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

| question | function | mzLib |
|---|---|---|
| which peptidoforms does a protein give, with which fragments? | `peptidoform_fragments()` | `Loaders.LoadUniprot`, `Protein.Digest`, `PeptideWithSetModifications.Fragment` |
| which modified peptides, and how many distinct backbones? | `digest_modified_peptides()`, `digest_distinct_base_sequences()` | the same digest |
| one ion series at a time | `digest_fragments_by_series()` | the same fragments |
| which annotated sites were not applied, and why? | `census_excluded()`, `census_explain()` | `ModificationCensus` |
| a peptide's m/z at a charge | `peptide_mz()` | none: the proton arithmetic is spelled out in `?peptide_mz` |
| a MetaMorpheus full sequence in Unimod accessions | `peptidoform_convert()` | `SequenceConversionService.Convert`, Unimod serializer |
| the same sequence in ProForma | `peptidoform_convert(target = "ProForma")` | `SequenceConversionService.Convert`, ProForma serializer |
| which sequences mzLib could not convert, and why | `records$status`, `records$incompatible_items` | `ConversionWarnings` |
| which notations mzLib can read and write | `source_formats`, `target_formats` | `SequenceConversionService.AvailableSourceFormats`, `AvailableTargetFormats` |

`peptidoform_fragments()` fetches a UniProt entry, applies its annotated modifications, digests
it and computes fragment ions for every peptidoform. The recording used here is the whole digest
of human serum albumin, P02768, at the defaults.

```{r}
digest <- peptidoform_fragments("P02768")
digest
```

## Three data.frames that join on `peptide_index`

```{r}
head(digest$peptides[, c("peptide_index", "base_sequence", "monoisotopic_mass",
                         "modification_count")])
head(digest$fragments)
digest$modifications
```

`monoisotopic_mass` and `neutral_mass` are **neutral** masses in daltons, not m/z.
`peptide_mz()` converts at a charge:

```{r}
head(peptide_mz(digest$peptides, charge = 2))
```

## Peptidoforms are not sequences

`peptides` has one row per sequence *and* modification placement, so its row count is larger
than the number of distinct sequences. Both are legitimate answers to "how many peptides", and
they are not interchangeable:

```{r}
nrow(digest$peptides)
digest_distinct_base_sequences(digest)
nrow(digest_modified_peptides(digest))
head(digest_modified_peptides(digest)[, c("full_sequence", "modification_count")], 3)
```

## Fragment series

ETD, the default, produces `c` and `z-dot` ions; HCD and CID produce `b` and `y`.

```{r}
digest_fragments_by_series(digest)
```

## Which modifications were applied, and which were not

mzLib loads only some UniProt feature types. The census says how many annotations were
applied, and why the rest were not:

```{r}
census_explain(digest)
head(census_excluded(digest))
```

## Before treating the list as complete

`max_isoforms` caps the peptidoforms generated per peptide, and the cap truncates silently.
`digest_truncated()` says whether it was reached:

```{r}
digest_truncated(digest)
```

Two defaults also decide which peptides exist at all: `min_length = 7` drops everything shorter,
and mzLib's `"trypsin|P"` *applies* the proline rule - the reverse of the MaxQuant and Mascot
spelling. `?peptidoform_fragments` gives the numbers.

## Converting full sequences to Unimod or ProForma

A MetaMorpheus result names each modification the way mzLib's databases do:
`[UniProt:N-acetylserine on S]`, `[Common Variable:Oxidation on M]`. Most other tools want a
Unimod accession instead, such as `[UNIMOD:1]`. `peptidoform_convert()` hands each sequence to
mzLib's `SequenceConversionService` and returns one row per input, in order. mzLibR maps no
modification itself; every output is mzLib's.

This example uses `BottomUpExample.psmtsv`, a MetaMorpheus search result that ships with mzLib's
tests. Read its full sequences, then convert them:

```{r}
psms <- readers_read_results("BottomUpExample.psmtsv")
unimod <- peptidoform_convert(psms$records$full_sequence)
unimod
head(unimod$records[, c("status", "output")], 5)
unimod$converted_count == unimod$record_count
```

UniProt's tele-methylhistidine became `UNIMOD:34` (Methyl). Carbamidomethyl became `UNIMOD:4`, and
oxidation became `UNIMOD:35`. Unmodified sequences pass through unchanged.

Every result lists the notations mzLib has registered, so you never need to guess a name:

```{r}
unimod$source_formats
unimod$target_formats
```

### ProForma does not resolve UniProt modifications yet

Ask for ProForma and the same file gives a different answer. Carbamidomethyl and oxidation become
UNIMOD accessions. The UniProt modification is written back under its mzLib name, and the row
still says `converted`:

```{r}
proforma <- peptidoform_convert(psms$records$full_sequence, target = "ProForma")
proforma$records$output[c(1, 5)]
proforma$records$status[1]
```

The cause is in mzLib, not in mzLibR (mzLib#1401). mzLib's ProForma serializer looks modifications
up only in MetaMorpheus's own list, which has no UniProt entries; the Unimod serializer looks them
up in every list mzLib loads. Until mzLib fixes this, **convert to Unimod when your sequences carry
UniProt modifications**, and in ProForma output treat any bracket that is not a `UNIMOD:` term as
unresolved. mzLibR does not patch around the gap: a patch here would leave the same gap in
pyMzLib, mzLibRust and MetaMorpheus. The `pro_forma` column that `readers_read_records()` gives a
`.psmtsv` comes from the same serializer, so it has the same gap.

### When mzLib cannot convert a sequence

What happens to a modification the target cannot write depends on `mode`, which is mzLib's
`SequenceConversionHandlingMode`:

| `mode` | the row | `output` |
|---|---|---|
| `"ReturnNull"` (default) | `failed` | `NA` |
| `"RemoveIncompatibleElements"` | `converted_with_warnings` | the sequence without that modification |
| `"UsePrimarySequence"` | `converted_with_warnings` | the sequence without that modification |
| `"ThrowException"` | no rows: the call raises an `mzlib_usage_error` naming the first such sequence | none |

`status` is mzLib's own verdict: `converted` (an output, and nothing recorded against it),
`converted_with_warnings` (an output, but mzLib noted something, such as a dropped modification or
a skipped character) or `failed` (no output).

These four sequences include one modification that has no Unimod accession:

```{r}
result <- peptidoform_convert(c(
  "[UniProt:N-acetylserine on S]SEQK",
  "PEPK[UniProt:N6,N6-dimethyllysine on K]R",
  "PEPM[Common Variable:Oxidation on M]K",
  "PEPK[Made Up:Not a modification on K]R"
))
result$records$output
bad <- result$records[result$records$status != "converted", ]
bad[, c("input", "status", "failure_reason")]
bad$incompatible_items[[1]]
```

`failure_reason` is `NA` here. Under `"ReturnNull"`, mzLib's Unimod serializer records the
incompatible modification but no reason code, so read `incompatible_items` to see what failed.

**Split ambiguous sequences first.** When MetaMorpheus cannot tell candidates apart, it joins their
full sequences with `|`. mzLib's parser does not refuse this (mzLib#1405): it skips each `|` with a
warning and joins the candidates into one sequence. The row is `converted_with_warnings`, but the
output is not a real peptide. Split on `|` before you convert.

`?peptidoform_convert` lists every parameter, field and caveat.

## What to cite

The methods and resources behind the functions on this page, from their specs:

<!-- BEGIN cite: generated by scripts/render-cite.R from the verb specs; do not edit -->

- [doi:10.1002/pmic.200300744](https://doi.org/10.1002/pmic.200300744): Unimod, the modification database whose accessions the Unimod target writes (Creasy and Cottrell 2004, Proteomics).
- [doi:10.1021/acs.jproteome.1c00771](https://doi.org/10.1021/acs.jproteome.1c00771): ProForma 2.0, the ProForma target's notation (LeDuc et al. 2022, J Proteome Res).
- [doi:10.1093/nar/gkae1010](https://doi.org/10.1093/nar/gkae1010): UniProt, the source of the entry, its sequence and its annotated modifications (UniProt Consortium 2025, NAR).

<!-- END cite -->

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
