---
title: "Reading mass-spectrometry and search-result files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading mass-spectrometry and search-result files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, made on mzLib's own
# test files, so the page builds anywhere. With a bridge installed, the same calls read your files.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

mzLib recognises dozens of file types: instrument data (mzML, Thermo `.raw`, Bruker and timsTOF
`.d`, MGF, msalign) and the output of a dozen search and deconvolution tools. The `readers_`
functions ask it what a file is and read its records into data.frames.

| question | function | mzLib |
|---|---|---|
| which file types does mzLib know, and what can each give me? | `readers_formats()` | `SupportedFileType` |
| what is this file? | `readers_identify()` | `SupportedFileTypeExtensions.ParseFileType` |
| what scans does this run hold, with peaks on request? | `readers_read_spectra()` | `MsDataFileReader.GetDataFile` |
| which peptides did the search find, ready to quantify? | `readers_read_results()` | `FileReader.ReadQuantifiableResultFile` |
| which deconvolved MS1 features? | `readers_read_features()` | `IMs1FeatureFile.GetMs1Features` |
| which spectral matches, with every engine score? | `readers_read_matches()` | `ISpectralMatch`, `MzIdentMLResultFile` |
| everything this file has, in its own fields | `readers_read_records()` | `IResultFile.LoadResults` |
| per-sample protein-group, peptide or PTM-site quantities | `readers_read_protein_groups()`, `readers_read_quantified_peptides()`, `readers_read_occupancy()` | `ProteinGroupFromTsvFile`, `QuantifiedPeptideFile`, `ModificationOccupancyCell` |
| the same, for many files in one call | the `_many` form of each | the same readers, `threads` at a time |

The one idea to take away: **mzLib does not read every format into one uniform shape.** The
formats fall into a few disjoint *views*, and many belong to none. This page shows how to tell
which you have, and which function reads it.

## What is this file?

```{r}
formats <- readers_formats()
nrow(formats)
formats[formats$is_quantifiable, c("file_type", "extension", "reader")]
```

A format's `views` say which cross-format projection it offers. An empty `views` is a real and
common answer: mzLib can parse the file, but no function other than `readers_read_records()`
reads it.

```{r}
table(vapply(formats$views, function(v) if (length(v)) paste(v, collapse = "+") else "(none)",
             character(1)))
```

For one file, `readers_identify()` resolves the type without parsing the contents:

```{r}
readers_identify("PXD078927_msgf_1_1_0.mzid")
```

## Five ways to read, and how to choose

| the file offers | read it with | columns |
|---|---|---|
| `spectra` | `readers_read_spectra()` | one row per scan |
| `quantifiable` | `readers_read_results()` | the uniform record view FlashLFQ takes |
| `ms1_features` | `readers_read_features()` | deconvolved MS1 features |
| `spectral_match` | `readers_read_matches()` | identifications, with q-values and scores for mzIdentML |
| anything | `readers_read_records()` | the format's own fields |

MetaMorpheus and FlashLFQ quantification tables have readers of their own, below.

### Spectra

Scan headers always; peak arrays only when asked for with `peaks = TRUE`, because they are
roughly the size of the file.

```{r}
scans <- readers_read_spectra("sliced_ethcd.mzML", limit = 3)
scans
scans$records[, c("one_based_scan_number", "ms_order", "retention_time", "peak_count",
                  "selected_ion_mz")]
```

With `peaks = TRUE`, `mz` and `intensity` are list columns, one numeric vector per scan:

```{r}
one <- readers_read_spectra("sliced_ethcd.mzML", peaks = TRUE, ms_order = 1, limit = 1)
c(peaks = length(one$records$mz[[1]]), peak_count = one$records$peak_count[[1]])
```

`retention_time` is in minutes for every spectra format. `selected_ion_mz` is `NA` on an MS1
scan, which has no precursor. `?readers_read_spectra` lists every column with its unit and what
`NA` means in it.

### Search results in the uniform view

```{r}
psms <- readers_read_results("FraggerPsm_FragPipev21.1_psm.tsv", limit = 2)
psms
psms$records[, c("base_sequence", "charge_state", "retention_time", "is_decoy")]
```

Read the `caveats` before trusting a column. Here they say that MSFragger records no decoy
label, so `is_decoy` is `NA` - unknown - rather than a `FALSE` that would read as "target".
Convert retention times with `readers_retention_time_in_minutes()` rather than by hand: it uses
the file's own `retention_time_unit`, and refuses when mzLib gives no basis for one.

```{r}
readers_retention_time_in_minutes(psms)
```

### Features, matches, and everything else

```{r}
features <- readers_read_features("Ms1Feature_TopFDv1.6.2_ms1.feature", limit = 5)
features$retention_time_unit
try(readers_retention_time_in_minutes(features))
```

TopFD changed the unit of `_ms1.feature` retention times at v1.7.0 without changing the format,
so the honest unit is `"unknown"`, and conversion refuses rather than guessing.

mzIdentML, which most search engines export, reads through the match view with its q-value,
rank and threshold - and with `scores = TRUE`, one row per match and engine score:

```{r}
matches <- readers_read_matches("PXD078927_msgf_1_1_0.mzid", limit = 3)
matches$records[, c("one_based_scan_number", "base_sequence", "q_value", "rank",
                    "pass_threshold")]
scored <- readers_read_matches("PXD078927_msgf_1_1_0.mzid", limit = 1, scores = TRUE)
scored$records[, c("match_index", "score_name", "score_value")]
```

Nothing in the match view is filtered: every candidate rank is a row, and a column a format has
no source for is `NA` in every row and named in `absent_fields`. Casanovo is de novo, so it has
no decoy label and no q-value:

```{r}
denovo <- readers_read_matches("Casanovo_5.0.0.mztab")
denovo$absent_fields
```

`readers_read_records()` reads any recognised file into that format's own fields, under mzLib's
names, so a column can be looked up in the mzLib source:

```{r}
prsms <- readers_read_records("ToppicPrsm_TopPICv1.6.2_prsm.tsv")
prsms$record_type
prsms$records[, c("one_based_scan_number", "base_sequence", "e_value")]
prsms$excluded_fields
```

`excluded_fields` names what could not become a column, with the reason, so a missing column is
never mistaken for a field the format does not have.

## Quantification tables

mzLib reads the tables MetaMorpheus and FlashLFQ write after quantifying. Each is wide on
disk, one column per sample, and arrives long: one row per record per sample. `limit` and
`offset` count records, so one group's samples are never split.

```{r}
groups <- readers_read_protein_groups("MetaMorpheus_1.1.11_AllQuantifiedProteinGroups.tsv",
                                      limit = 1)
groups
head(groups$records[, c("protein_group_name", "q_value", "sample_label", "intensity")])
```

The table is unfiltered - decoys, contaminants and groups above 1% FDR are all rows - and a
blank intensity is `NA`, not zero. `readers_read_quantified_peptides()` reads the peptide table,
where a FlashLFQ intensity of 0 is *not* a measurement and `detection_type` says why; and
`readers_read_occupancy()` reads the per-site PTM occupancy, one row per group, sample, basis
and site:

```{r}
sites <- readers_read_occupancy("MetaMorpheus_1.1.11_AllQuantifiedProteinGroups.tsv")
head(sites$records[sites$records$basis == "count",
                   c("protein_group_name", "position", "modification", "numerator",
                     "denominator")])
```

### RNA searches, and searches without quantification

The same three functions read two kinds of table they used to refuse:

- **An RNA search** writes `AllQuantifiedTranscriptGroups.tsv` and `AllQuantifiedOligos.tsv`
  (mzLib #1388). They are their own file types, but mzLib reads them with subclasses of the
  protein-group and peptide readers, so the columns keep their protein names:
  `protein_group_name` holds the transcript group, `sequence` the oligonucleotide, and
  `readers_read_occupancy()` returns RNA modifications such as `2'-O-methyluridine on U`.
- **A search with label-free quantification off** writes `AllProteinGroups.tsv`, and each file's
  `<file>_ProteinGroups.tsv` (mzLib #1365). MetaMorpheus writes `Intensity_` columns only when it
  quantified, so these rows have `spectral_count` and no `intensity`: the field is named in
  `absent_fields`, which means *no basis*, not zero.

On mzLib's own RNA test table:

```{r}
rna <- readers_read_protein_groups("MetaMorpheus_RNA_AllQuantifiedTranscriptGroups.tsv")
c(groups = rna$record_count, rows = nrow(rna$records))
fluc <- rna$records[rna$records$protein_group_name == "FLuc" & rna$records$sample_label == "1:1_1", ]
fluc[, c("protein_group_name", "sample_label", "spectral_count", "intensity")]
```

## Many files in one call

Every reader has a `_many` form that hands the whole list to one bridge process, which reads
`threads` files at once and returns one long table - the same table at any thread count. Never
loop the one-file function instead: each call starts a new process.

```{r}
runs <- readers_read_spectra_many(c("sliced_ethcd.mzML", "no-such-run.mzML", "withZeros.mgf"),
                                  threads = 2, on_error = "skip")
runs
runs$files[, c("file_type", "scan_count", "instrument_model", "error_kind")]
table(runs$records$source_index)
```

`records` starts with `source_index` - 1-based, so it indexes `files` - and `source_path`.
`files` has one row per input with everything a one-file read reports about it, including the
run's instrument and acquisition time for a spectra file; with `on_error = "skip"`, an input that
could not be read is a row with its error rather than a failed call.

## Large files

Every call starts one bridge process and reads the whole file, so `limit` and `offset` are a
window, not a cursor: paging re-reads the file each time. For a large file pass `out` to write a
tab-separated table and read it with `read.delim()`.

## What to cite

The methods and resources behind the functions on this page, from their specs:

<!-- BEGIN cite: generated by scripts/render-cite.R from the verb specs; do not edit -->

- No publication: these verbs cite none in their specs.

<!-- END cite -->

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
