---
title: "SDRF-Proteomics experimental designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SDRF-Proteomics experimental designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, so the page builds
# anywhere. With a bridge installed, the same calls read your files.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

Every other reader in mzLibR answers *what did the search find*. An SDRF-Proteomics file
answers *what was searched*: which sample, organism part, replicate and instrument settings.
That is the half you need to group results across experiments.

| question | function | mzLib |
|---|---|---|
| what does this SDRF say, row by row? | `sdrf_read()` | `SdrfDocument.LoadResults` |
| several SDRFs as one analysis table, with provenance | `sdrf_pool()` | `SdrfCollection.Merge` |
| is it well-formed? | `sdrf_validate()`, `sdrf_validate_many()` | `SdrfValidator.Validate` |
| does it describe its samples, or is it a valid skeleton? | `sdrf_assess()`, `sdrf_assess_many()` | `SdrfSampleInformativeness.Assess` |
| each sample's characteristics, with ages in years | `sdrf_samples()`, `sdrf_parse_ages()` | `SdrfSampleBlock.BySourceName`, `SdrfAge.TryParse` |
| do several SDRFs spell their terms alike? | `sdrf_lint()` | `SdrfDriftLint.Analyze` |
| can it drive a label-free quantification, and with what design? | `sdrf_design()` | `SdrfLabelFreeDesign.Read` (#1363) |

## Read one document

```{r}
doc <- sdrf_read("PXD000070.sdrf.tsv")
doc
```

An SDRF document is **row-major**, not a data.frame, for two reasons that are both real:

- **Column names repeat.** `comment[modification parameters]` legitimately appears once per
  modification, so a name identifies a *position*, not a column.
- **Rows can be ragged.** A row may carry fewer cells than the header has names, and mzLib keeps
  that so the file round-trips.

So `columns` is a character vector that may repeat, `rows` a list of character vectors, and the
accessors do the looking up:

```{r}
sdrf_value(doc, "characteristics[organism part]")
sdrf_indexes_of(doc, "comment[modification parameters]")
sdrf_all(doc, "comment[modification parameters]")[[1]]
```

`NA` from `sdrf_value()` means the document has no such column, or the row is too short to
reach it. The reserved words `"not available"` and `"not applicable"` are different: they are
values the experiment chose to write, and they come back as themselves.

When a data.frame is what you want, `sdrf_records()` gives one - lossy when a name repeats,
which `sdrf_has_repeated_columns()` tells you first:

```{r}
sdrf_has_repeated_columns(doc)
head(sdrf_records(doc)[, c("source name", "characteristics[organism]")])
```

A ragged document, from mzLib's own fixtures:

```{r}
ragged <- sdrf_read("PXD059974.sdrf.tsv")
sdrf_ragged_row_count(ragged)
```

## Pool several into one analysis table

Name the documents: the names become each row's provenance label.

```{r}
pooled <- sdrf_pool(c(malaria = "PXD000070.sdrf.tsv", colon = "PXD026824.sdrf.tsv"), limit = 4)
pooled
sdrf_source_documents(pooled)
```

Columns are the union over the documents, and a cell a document did not have is filled with
`"not available"` - so a fill rate computed on the pooled table is not the fill rate of the
originals. The pooled table is for analysis, not for deposit: two experiments may both have a
"Sample 1", so keep `sdrf_source_documents()` in any key.

## Is the document well-formed?

`sdrf_validate()` checks the specification's structural rules - required columns, ragged rows,
the uniqueness of `source name` + `assay name` + `comment[label]` - and locates each finding by
line:

```{r}
skeleton <- sdrf_validate("sdrf_skeleton.sdrf.tsv")
skeleton
skeleton$records[, c("severity", "rule", "line_number", "column_name")]
```

Warnings never make a document invalid. A document-wide finding, such as a missing column, has
`NA` for its line.

## Does it describe its samples?

A document can be perfectly valid and still say nothing: reserved words in every sample column,
one replicate number, no factor. `sdrf_assess()` is the gate for that, and its verdict is how
many of three checks pass:

```{r}
assessment <- sdrf_assess("sdrf_cohort.sdrf.tsv")
assessment
assessment$records[, c("role", "column_name", "filled", "distinct_values", "fill_rate")]
```

A real deposit, PXD000070, is `Partial`: it describes its samples and varies its replicates, but
no factor value varies, so it cannot say what was compared. A valid skeleton says nothing:

```{r}
pxd70 <- sdrf_assess("PXD000070.sdrf.tsv")
str(pxd70[c("verdict", "factor_value_varies", "sample_is_described", "biological_replicate_varies")])
empty <- sdrf_assess("sdrf_skeleton.sdrf.tsv")
empty$verdict
empty$records[, c("column_name", "filled", "rows")]
```

For a corpus, `sdrf_assess_many()` reads every document in one bridge call:

```{r}
batch <- sdrf_assess_many(c("sdrf_cohort.sdrf.tsv", "sdrf_skeleton.sdrf.tsv", "PXD000070.sdrf.tsv"))
batch$files[, c("path", "verdict")]
```

## Each sample, merged over its rows, with its age

`sdrf_samples()` gives one row per sample x column. Where a sample's rows disagree about a
column, mzLib withholds it rather than let the first row win, and `status` says
`"conflicting"`. Ages are parsed into years on the `characteristics[age]` rows:

```{r}
samples <- sdrf_samples("sdrf_cohort.sdrf.tsv")
samples$records[samples$records$status == "conflicting", c("source_name", "column_name")]
ages <- samples$records[samples$records$column_name == "characteristics[age]", ]
ages[, c("source_name", "value", "age_years", "age_precision", "age_refusal")]
```

A bare `63` is refused as `"no_unit"`: 63 years and 63 days are both plausible in one study.
`sdrf_parse_ages()` applies the same rules to any cells you have:

```{r}
sdrf_parse_ages(c("58Y", "40Y-85Y", ">=90Y", "6-8 weeks", "63"))$records[
  , c("cell", "years", "min_years", "max_years", "refusal")]
```

## Do several documents agree?

Every document can pass `sdrf_validate()` and a pooled table still be unusable because one writes
`"Homo sapiens"` and another `"homo sapiens"`. `sdrf_lint()` finds those, one row per spelling:

```{r}
drift <- sdrf_lint(c(cohort = "sdrf_cohort.sdrf.tsv", partner = "sdrf_cohort_partner.sdrf.tsv"))
drift
drift$records[, c("finding_index", "kind", "value", "occurrences")]
```

## Turn an SDRF into a quantification design

FlashLFQ and MetaMorpheus need an experimental design: which runs are the same condition, which
are biological replicates of one another, which are fractions of one sample. The SDRF already
says all of that. A hand-written `ExperimentalDesign.tsv` is how most people still give it, and
an invalid one is worse than none: MetaMorpheus skips quantification with one warning and no
error.

`sdrf_design()` reads the design out of the SDRF, or refuses and says why. It calls mzLib's
`SdrfLabelFreeDesign`, which runs every check MetaMorpheus's own design validator runs and reports
*all* the failures at once:

| question | function | mzLib |
|---|---|---|
| the design, or every reason there is none | `sdrf_design(path, condition_columns)` | `SdrfLabelFreeDesign.Read` (#1363) |
| only the runs one search read | `sdrf_design(..., searched_files = ...)` | `SdrfLabelFreeDesignOptions.SearchedFiles` |
| it as `flashlfq_quantify(spectra = )` | `sdrf_design_spectra()` | `SdrfLabelFreeDesign.Files` |
| it as `flashlfq_median_polish(design = )` | `sdrf_design_run_design()` | `SdrfLabelFreeDesign.Files` |
| MetaMorpheus's `ExperimentalDesign.tsv` | `sdrf_design(..., out = "ExperimentalDesign.tsv")` | `WriteExperimentalDesignTsv` |

The examples use mzLib's own design files: PXD067622, a TurboID proximity-labelling study (two
genotypes by four treatments, three biological replicates each), and PXD049018 (two pulldowns,
ten SDS-PAGE bands each).

### A valid design

PXD067622 has two factor columns. Say which make up the condition; their values are joined with
`_`, in the order you give:

```{r}
both <- c("factor value[genotype]", "factor value[treatment]")
d <- sdrf_design("PXD067622.sdrf.tsv", condition_columns = both)
d
head(d$records[, c("file_name", "condition", "biological_replicate")], 3)
```

`biological_replicate`, `technical_replicate` and `fraction` are **0-based** - mzLib's
`SpectraFileInfo` coordinates, exactly what the FlashLFQ functions take - although every other
position in mzLibR is 1-based. `out` writes MetaMorpheus's file, which starts at 1; mzLib adds the
one when it writes, and nowhere else. Never add it yourself.

Hand the design straight to FlashLFQ:

```{r}
head(sdrf_design_spectra(d), 2)
```

The SDRF names files without a directory. When your spectra live elsewhere, or the search read
only some of them, pass `searched_files`: each must be named **exactly** (case and extension) by
one row, rows for other files are dropped and listed in `notes`, and your paths become
`full_path`.

### A refusal lists every reason

PXD049018's depositors did not say which pulldown was treated: its `factor value[treatment]` is
`not available` on every row. A condition built from an unknown factor would pair samples nobody
said were alike, so mzLib refuses, once per row, naming each file:

```{r}
refused <- sdrf_design("PXD049018.sdrf.tsv", condition_columns = both)
refused$is_valid
c(refusals = length(refused$refusals), files = refused$file_count)
writeLines(refused$refusals[[1]])
# The hand-offs refuse a refused design, naming every reason:
inherits(try(sdrf_design_spectra(refused), silent = TRUE), "try-error")
```

**A refusal is a result, not an error.** `sdrf_design()` returns normally with `is_valid` `FALSE`;
only the two hand-offs stop, so a refused design cannot reach FlashLFQ by accident. Without
`condition_columns`, mzLib uses the document's only factor column - and PXD067622 has two, so
leaving them undeclared is itself a refusal: joining every factor could split a condition on a
nuisance factor.

### Replicate numbers are ranked within each condition

A drafted SDRF often copies the study-wide run index out of the file names. MetaMorpheus wants
replicates numbered within each condition, so mzLib ranks them back, and records every
renumbering in `notes`, the only place it is kept:

```{r}
ranked <- sdrf_design("PXD067622_studywide.sdrf.tsv", condition_columns = both)
writeLines(ranked$notes[[2]])
identical(ranked$records, d$records)
```

**Label-free only.** An isobaric SDRF needs a channel design that `sdrf_design()` does not build.
For the channels themselves, see the isobaric kits article.

## Instead of `readers_read_records()`

`readers_read_records()` recognises SDRF too, but joins each row into one semicolon-separated
string, and SDRF's own `NT=...;AC=...` grammar puts semicolons inside cells. Use `sdrf_read()`.

## What to cite

The methods and resources behind the functions on this page, from their specs:

<!-- BEGIN cite: generated by scripts/render-cite.R from the verb specs; do not edit -->

- [doi:10.1038/s41467-021-26111-3](https://doi.org/10.1038/s41467-021-26111-3): SDRF-Proteomics sample columns (characteristics, factor value) (Dai et al., Nat Commun 12, 5854, 2021).

<!-- END cite -->

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
