---
title: "Label-free quantification with FlashLFQ"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a small recording of the real bridge, so the page
# builds anywhere. With a bridge installed, the same call quantifies your runs.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

`flashlfq_quantify()` takes a search's identifications and the mzML runs they came from, and
returns FlashLFQ's peptide, protein and peak tables. The recording used here is deliberately
tiny - two runs, two peptides - so every table fits on the page.

```{r}
quant <- flashlfq_quantify("AllPSMs.psmtsv", c("run_3.mzML", "run_4.mzML"),
                           match_between_runs = TRUE)
quant
```

## Four tidy tables

```{r}
quant$spectra_files[, c("file_name", "peak_count", "mbr_peak_count")]
quant$peptides
quant$proteins
quant$peaks
```

`peptides` and `proteins` are long: one row per peptide or protein group per run, with the run
in `file_name`. Intensities are in the instrument's units; `retention_time` in `peaks` is the
apex, in minutes.

## 0 and `NA` mean different things

A **peptide** intensity of `0` means the peptide was not measured in that run. A **protein**
intensity of `NA` means FlashLFQ could not resolve a number at all. mzLibR keeps the two apart,
so `mean()` over a protein column returns `NA` rather than a confidently wrong number, and
`na.rm = TRUE` is a choice you make visibly.

## Count transfers from the peaks

With `match_between_runs = TRUE`, count transfers from `peaks`, never from the peptide roll-up,
which drops most of them. On mzLib's own K562 pair the peaks hold 140 true transfers and the
peptide table shows 52.

```{r}
flashlfq_mbr_peaks(quant)
flashlfq_mbr_peak_count(quant)
flashlfq_mbr_rescued_peptide_count(quant)
```

## Re-quantify proteins under a new design

`flashlfq_median_polish()` reruns only the protein roll-up, from the `QuantifiedPeptides.tsv`
FlashLFQ wrote, without re-reading any spectra. The design says which runs are replicates of which
sample, and the result is keyed by sample:

```{r}
polished <- flashlfq_median_polish("QuantifiedPeptides.tsv",
  design = data.frame(file_name = c("run_3", "run_4"), condition = c("control", "treated"),
                      biological_replicate = 0))
polished$samples
polished$proteins
```

## Before a real run

- Use a MetaMorpheus `.psmtsv` or `.osmtsv`, and mzML runs; convert `.raw` first.
- Give `spectra` a `condition` and `biological_replicate` so FlashLFQ knows which runs are
  comparable; match-between-runs needs a balanced design.
- `max_threads` defaults to 1 here and to every core in pyMzLib and mzLibRust;
  `?flashlfq_quantify` says why.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
