---
title: "Digesting and fragmenting peptidoforms"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge fetching UniProt, so
# the page builds offline. With a bridge installed, the same calls fetch the entry live.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

`peptidoform_fragments()` fetches a UniProt entry, applies its annotated modifications, digests
it and computes fragment ions for every peptidoform. The recording used here is human serum
albumin, P02768, with at most one modification per peptide.

```{r}
digest <- peptidoform_fragments("P02768", max_modifications = 1)
digest
```

## Three data.frames that join on `peptide_index`

```{r}
head(digest$peptides[, c("peptide_index", "base_sequence", "monoisotopic_mass",
                         "modification_count")])
head(digest$fragments)
digest$modifications
```

`monoisotopic_mass` and `neutral_mass` are **neutral** masses in daltons, not m/z.
`peptide_mz()` converts at a charge:

```{r}
peptide_mz(digest$peptides, charge = 2)
```

## Peptidoforms are not sequences

`peptides` has one row per sequence *and* modification placement, so its row count is larger
than the number of distinct sequences. Both are legitimate answers to "how many peptides", and
they are not interchangeable:

```{r}
nrow(digest$peptides)
digest_distinct_base_sequences(digest)
digest_modified_peptides(digest)[, c("full_sequence", "modification_count")]
```

## Fragment series

ETD, the default, produces `c` and `z-dot` ions; HCD and CID produce `b` and `y`.

```{r}
digest_fragments_by_series(digest)
```

## Which modifications were applied, and which were not

mzLib loads only some UniProt feature types. The census says how many annotations were
applied, and why the rest were not:

```{r}
census_explain(digest)
head(census_excluded(digest))
```

## Before treating the list as complete

`max_isoforms` caps the peptidoforms generated per peptide, and the cap truncates silently.
`digest_truncated()` says whether it was reached:

```{r}
digest_truncated(digest)
```

Two defaults also decide which peptides exist at all: `min_length = 7` drops everything shorter,
and mzLib's `"trypsin|P"` *applies* the proline rule - the reverse of the MaxQuant and Mascot
spelling. `?peptidoform_fragments` gives the numbers.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
