---
title: "Finding and downloading PRIDE Archive data"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# The listings on this page are answered from recordings of the real bridge talking to PRIDE,
# so the page builds offline. With a bridge installed, the same calls query EBI.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

The `pride_` functions list a PRIDE Archive project's files and download the ones you choose,
using mzLib's paging and URL resolution.

## Find a project

`pride_search()` is the discovery step: it turns a subject into the accessions every other
function takes, with every page fetched and no accession repeated.

```{r}
hits <- pride_search("plasmodium falciparum schizont")
hits[, c("accession", "submission_type", "publication_date")]
hits$organisms[[1]]
hits$matched_fields[[1]]
```

A hit is a search-index projection, not the project's metadata: vocabulary fields arrive as
display strings with no accessions, a count of 0 means "not reported", and `project_file_names`
is not the manifest. Follow the `accession`.

## List, then choose with `[`

```{r}
files <- pride_list_files("PXD000001")
files[, c("file_name", "category", "size_mb", "downloadable")]
```

Choose rows with ordinary subsetting, then pass them to `pride_download_files()`. That says
things the `category` and `extensions` filters of `pride_download()` cannot, such as "under
5 MB" or "everything except the raw file".

```{r}
small <- files[files$size_mb < 5 & files$downloadable, ]
small$file_name
pride_total_size_bytes(small)
```

```{r, eval = FALSE}
pride_download_files(small, "downloads")   # needs the network
```

## Two things the manifest will not tell you

**Sizes can be the decompressed size.** For some compressed files PRIDE reports the size of the
content, not of the download: the `.mgf.gz` above is listed at 16 MB and transfers 6 MB. Treat
`pride_total_size_bytes()` as an upper bound.

**The manifest can be incomplete.** PRIDE's REST API publishes 8 files for PXD000001; the FTP tree
holds 13, including the two largest. `pride_list_ftp_files()` walks the tree instead:

```{r}
listing <- pride_list_ftp_files("PXD000001")
listing[, c("relative_path", "approximate_size_mb")]
pride_approximate_total_size_bytes(listing)
```

A file that appears only there is fetched from its `url` with any HTTPS client, for example
`download.file()`.

## Filters that match nothing are errors

`pride_download()` with a `category` or `extensions` filter that matches nothing raises, rather
than reporting success with nothing downloaded. A compressed file's extension is `.gz`, so
`extensions = ".mgf"` matches nothing in PXD000001:

```{r, eval = FALSE}
pride_download("PXD000001", "downloads", category = "PEAK", extensions = ".gz")
```

An unknown accession raises `mzlib_project_not_found`, and an EBI outage raises
`mzlib_service_unavailable`, which is the one to retry. See `?mzlib_error`.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
