---
title: "Protein databases: organisms, genes, unique peptides and GO"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, made on the small
# databases in pyMzLib's test fixtures, so the page builds anywhere. With a bridge installed, the
# same calls read your databases.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

A search result names proteins by accession. The questions it cannot answer on its own are
answered by mzLib from the protein database you searched:

| question | function | mzLib |
|---|---|---|
| what is this accession: organism, taxon, genes, GO terms? | `proteins_read()` | `ProteinDbLoader` |
| which Ensembl gene is it, reproducibly? | `proteins_resolve_genes()` | `EnsemblGeneResolver` |
| does this peptide identify one protein? | `proteins_classify_peptides()` | `PeptideUniquenessClassifier` |
| which GO terms does each protein group carry, every member kept? | `proteins_annotate_go()` | `GoGroupAnnotator` (#1353), `ToGoAnnotationGroups` (#1366) |
| where do I get a go.obo, and keep it pinned? | `proteins_update_go()` | `Loaders.UpdateGeneOntology` (#1353) |

The first three take one database or several - UniProt XML or FASTA, optionally gzipped - and
read several in **one** bridge call, `threads` at a time. Mark contaminant databases with
`contaminants`: it changes answers.

## What is this accession?

```{r}
db <- proteins_read(c("human_subset.xml", "human_extra.fasta", "mouse_aifm1.fasta"),
                    contaminants = "contaminants.fasta",
                    tables = c("proteins", "go_terms", "ensembl_genes"))
db
db$proteins[, c("accession", "organism", "ncbi_taxonomy_id", "primary_gene_name",
                "length", "monoisotopic_mass", "is_contaminant")]
```

`length` is in residues and `monoisotopic_mass` in daltons, of the unmodified sequence as written
plus one water - the precursor, not the mature protein. `ncbi_taxonomy_id` is kept a string: it
is an identifier, not a quantity.

GO terms and Ensembl gene links are long tables, one row per protein and term or transcript,
returned only when asked for:

```{r}
head(db$go_terms[, c("accession", "go_id", "aspect", "term_name")])
db$ensembl_genes[, c("accession", "gene_id", "versioned_gene_id", "gene_version")]
```

### A FASTA cannot say

A FASTA header carries organism, taxon and gene name, and nothing else. So a FASTA contributes no
GO or Ensembl rows - and says so, rather than leaving the tables silently short:

```{r}
db$files[, c("file_type", "contaminant", "protein_count")]
db$files$absent_fields
```

Empty there means "this format is silent", never "this protein has none". Read the UniProt XML
of the same proteome for those.

`source_index` in every table is 1-based, so it indexes `files` directly:

```{r}
table(basename(db$files$path[db$proteins$source_index]))
```

## Which gene is it, reproducibly?

`proteins_resolve_genes()` resolves each protein to stable Ensembl gene ids, **counted against a
gene set you supply**: an Ensembl GTF for one release. Nothing is downloaded or defaulted, because
the release is part of the answer.

```{r}
genes <- proteins_resolve_genes(c("human_subset.xml", "human_extra.fasta"),
                                contaminants = "contaminants.fasta",
                                gtf = "Homo_sapiens.GRCh38.116.gtf",
                                xref = "Homo_sapiens.GRCh38.116.uniprot.tsv")
genes$gene_set[c("source_file_name", "release", "genome_build")]
genes$outcome_counts
genes$resolutions[, c("accession", "outcome", "n_genes", "gene_id", "gene_symbol",
                      "off_primary_genes", "ensembl_xref_agrees")]
```

Every protein gets exactly one outcome. A `multi_gene` protein has one row per gene, never a
pick. `off_primary_only` means its genes exist only on ALT haplotypes or patches outside the gene
set - use Ensembl's primary-assembly GTF. `ensembl_xref_agrees` is `NA` when there is nothing to
compare, which is unknown, not false. Keep `genes$gene_set$sha256` with your results: it pins the
exact gene set.

## Does this peptide identify one protein?

```{r}
calls <- proteins_classify_peptides(
  c("VGVNGFGR", "LVLNGNPLTLFQER", "ALSEQINIFFDYSGR", "YLYEIAR", "AEFVEVTK", "PEPTIDEK"),
  c("human_subset.xml", "human_extra.fasta"),
  contaminants = "contaminants.fasta")
calls$peptides[, c("peptide", "sharing", "accession_count")]
calls$sharing_counts
```

- `LVLNGNPLTLFQER` is GAPDH's `LVINGNPITIFQER`: **I and L are one residue**, because a mass
  spectrometer cannot tell them apart.
- `ALSEQINIFFDYSGR` is in three isoforms of one gene: `SharedWithinGene` supports the gene, not
  an isoform.
- `YLYEIAR` is in human and bovine albumin, which share no gene. Leave the contaminant database
  out and it would read as `Unique`.

```{r}
calls$peptides$accessions[[4]]
calls$peptides$shared_gene_keys[[3]]
```

Containment, not digestion: a protein contains a peptide if its sequence contains it anywhere,
whatever the protease. That is deliberately conservative, so a peptide called `Unique` cannot be
explained by another entry at a site the search's cleavage rules skipped.

## Which GO terms does each protein group carry?

A search reports *protein groups*, and a group can have several members: two histone H2A.Z
variants no peptide could tell apart are one group, `P0C0S5|Q71UI9`. Taking the first accession
throws away half the evidence, and it is not even a choice MetaMorpheus made: it sorts members and
never picks a leading protein. And UniProt records the most specific terms, so a protein annotated
to *mitochondrial inner membrane* is not, as written, "in the mitochondrion" at all.

`proteins_annotate_go()` answers both, with mzLib's `GoGroupAnnotator` reading the protein-group
table MetaMorpheus wrote: **one row per (group, term)** that *any* member holds, directly or
through an ancestor (`is_a` and `part_of`), and every row says which members carry it. The data
here is real: mzLib's copy of a MetaMorpheus 1.1.11 protein-group table from PXD036557, the
UniProt entries it names, and GO release 2026-07-26, trimmed to the terms those proteins reach.

### First, get a go.obo - once, and keep it

Terms are added, renamed and moved between GO releases, so a GO result means something only
relative to one release. `proteins_annotate_go()` never downloads anything; fetching is its own
function, called on purpose:

```{r}
update <- proteins_update_go("go.obo")
update$go[c("release", "term_count")]
```

Call it again later and a newer release replaces the file, while the old one is kept beside it
as `go.obo.<timestamp>`, so a result you already published can still be reproduced.

### Annotate the groups

```{r}
go <- proteins_annotate_go("PXD036557_AllQuantifiedProteinGroups.tsv", "pxd036557_proteins.xml",
                           go_obo = "go-pxd036557.obo", category_map = "organelle_map.tsv")
go
c(table_rows = go$table_row_count, decoys = go$decoy_group_count, groups = go$group_count,
  rows = go$row_count)
```

The decoy group is skipped, because decoys carry no GO. Albumin, a contaminant, gets one row with
no term, and its `annotation_status` says why:

```{r}
go$annotations[go$annotations$protein_group == "P02769",
               c("protein_group", "go_id", "annotation_status", "n_with", "n_members")]
```

Both histones are annotated to the nucleosome, each on its own evidence:

```{r}
histones <- go$annotations[go$annotations$protein_group == "P0C0S5|Q71UI9", ]
nucleosome <- histones[histones$go_id %in% "GO:0000786", ]
nucleosome[, c("go_name", "aspect", "n_with", "n_members", "propagated", "inherited")]
nucleosome$accession_used[[1]]
nucleosome$evidence_by_member[[1]]
```

### Union, consensus, direct only: filters, not modes

The table is the **union** over a group's members. Everything narrower is one comparison on the
rows:

```{r}
c(union = nrow(histones), consensus = sum(histones$n_with == histones$n_members),
  direct = sum(!histones$propagated))
histones$accession_used[histones$go_name %in% "euchromatin"]
```

*Euchromatin* is one histone's term, not the other's. For an enrichment tool that propagates on
its own, keep the rows with `propagated` false.

### In your own vocabulary

`category_map` maps terms to the categories you care about, through anchors: a term belongs to a
category when the anchor is the term or one of its ancestors. mzLib ships no vocabulary.

```{r}
head(go$categories$records)
```

### Pin what you used

Keep `go$go$sha256` and `go$annotation_database$sha256` with your results. The header's counters
count **groups at q <= 0.01**, not rows; the rows themselves are never filtered on q.

```{r}
go$go[c("release", "sha256")]
go$header[c("counter_q_value_max", "status_annotated", "status_contaminant")]
```

When the UniProt XML is newer than the go.obo it cites terms the release lacks, and the call fails
naming every one. Fetch a newer go.obo, or pass `skip_unknown_go_ids = TRUE` to drop those ids and
have them listed in `unresolved_go_ids`.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
