---
title: "Protein databases: organisms, genes and unique peptides"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, made on the small
# databases in pyMzLib's test fixtures, so the page builds anywhere. With a bridge installed, the
# same calls read your databases.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

A search result names proteins by accession. Three questions it cannot answer on its own are
answered by mzLib from the protein database you searched:

| question | function |
|---|---|
| what is this accession: organism, taxon, genes, GO terms? | `proteins_read()` |
| which Ensembl gene is it, reproducibly? | `proteins_resolve_genes()` |
| does this peptide identify one protein? | `proteins_classify_peptides()` |

All three take one database or several - UniProt XML or FASTA, optionally gzipped - and read
several in **one** bridge call, `threads` at a time. Mark contaminant databases with
`contaminants`: it changes answers.

## What is this accession?

```{r}
db <- proteins_read(c("human_subset.xml", "human_extra.fasta", "mouse_aifm1.fasta"),
                    contaminants = "contaminants.fasta",
                    tables = c("proteins", "go_terms", "ensembl_genes"))
db
db$proteins[, c("accession", "organism", "ncbi_taxonomy_id", "primary_gene_name",
                "length", "monoisotopic_mass", "is_contaminant")]
```

`length` is in residues and `monoisotopic_mass` in daltons, of the unmodified sequence as written
plus one water - the precursor, not the mature protein. `ncbi_taxonomy_id` is kept a string: it
is an identifier, not a quantity.

GO terms and Ensembl gene links are long tables, one row per protein and term or transcript,
returned only when asked for:

```{r}
head(db$go_terms[, c("accession", "go_id", "aspect", "term_name")])
db$ensembl_genes[, c("accession", "gene_id", "versioned_gene_id", "gene_version")]
```

### A FASTA cannot say

A FASTA header carries organism, taxon and gene name, and nothing else. So a FASTA contributes no
GO or Ensembl rows - and says so, rather than leaving the tables silently short:

```{r}
db$files[, c("file_type", "contaminant", "protein_count")]
db$files$absent_fields
```

Empty there means "this format is silent", never "this protein has none". Read the UniProt XML
of the same proteome for those.

`source_index` in every table is 1-based, so it indexes `files` directly:

```{r}
table(basename(db$files$path[db$proteins$source_index]))
```

## Which gene is it, reproducibly?

`proteins_resolve_genes()` resolves each protein to stable Ensembl gene ids, **counted against a
gene set you supply**: an Ensembl GTF for one release. Nothing is downloaded or defaulted, because
the release is part of the answer.

```{r}
genes <- proteins_resolve_genes(c("human_subset.xml", "human_extra.fasta"),
                                contaminants = "contaminants.fasta",
                                gtf = "Homo_sapiens.GRCh38.116.gtf",
                                xref = "Homo_sapiens.GRCh38.116.uniprot.tsv")
genes$gene_set[c("source_file_name", "release", "genome_build")]
genes$outcome_counts
genes$resolutions[, c("accession", "outcome", "n_genes", "gene_id", "gene_symbol",
                      "off_primary_genes", "ensembl_xref_agrees")]
```

Every protein gets exactly one outcome. A `multi_gene` protein has one row per gene, never a
pick. `off_primary_only` means its genes exist only on ALT haplotypes or patches outside the gene
set - use Ensembl's primary-assembly GTF. `ensembl_xref_agrees` is `NA` when there is nothing to
compare, which is unknown, not false. Keep `genes$gene_set$sha256` with your results: it pins the
exact gene set.

## Does this peptide identify one protein?

```{r}
calls <- proteins_classify_peptides(
  c("VGVNGFGR", "LVLNGNPLTLFQER", "ALSEQINIFFDYSGR", "YLYEIAR", "AEFVEVTK", "PEPTIDEK"),
  c("human_subset.xml", "human_extra.fasta"),
  contaminants = "contaminants.fasta")
calls$peptides[, c("peptide", "sharing", "accession_count")]
calls$sharing_counts
```

- `LVLNGNPLTLFQER` is GAPDH's `LVINGNPITIFQER`: **I and L are one residue**, because a mass
  spectrometer cannot tell them apart.
- `ALSEQINIFFDYSGR` is in three isoforms of one gene: `SharedWithinGene` supports the gene, not
  an isoform.
- `YLYEIAR` is in human and bovine albumin, which share no gene. Leave the contaminant database
  out and it would read as `Unique`.

```{r}
calls$peptides$accessions[[4]]
calls$peptides$shared_gene_keys[[3]]
```

Containment, not digestion: a protein contains a peptide if its sequence contains it anywhere,
whatever the protease. That is deliberately conservative, so a peptide called `Unique` cannot be
explained by another entry at a site the search's cleavage rules skipped.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
