---
title: "SDRF-Proteomics experimental designs"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(mzLibR)
# Every call on this page is answered from a recording of the real bridge, so the page builds
# anywhere. With a bridge installed, the same calls read your files.
.mzlibr_example <- mzLibR:::replay_bridge_start()
```

Every other reader in mzLibR answers *what did the search find*. An SDRF-Proteomics file
answers *what was searched*: which sample, organism part, replicate and instrument settings.
That is the half you need to group results across experiments.

## Read one document

```{r}
doc <- sdrf_read("PXD000070.sdrf.tsv")
doc
```

An SDRF document is **row-major**, not a data.frame, for two reasons that are both real:

- **Column names repeat.** `comment[modification parameters]` legitimately appears once per
  modification, so a name identifies a *position*, not a column.
- **Rows can be ragged.** A row may carry fewer cells than the header has names, and mzLib keeps
  that so the file round-trips.

So `columns` is a character vector that may repeat, `rows` a list of character vectors, and the
accessors do the looking up:

```{r}
sdrf_value(doc, "characteristics[organism part]")
sdrf_indexes_of(doc, "comment[modification parameters]")
sdrf_all(doc, "comment[modification parameters]")[[1]]
```

`NA` from `sdrf_value()` means the document has no such column, or the row is too short to
reach it. The reserved words `"not available"` and `"not applicable"` are different: they are
values the experiment chose to write, and they come back as themselves.

When a data.frame is what you want, `sdrf_records()` gives one - lossy when a name repeats,
which `sdrf_has_repeated_columns()` tells you first:

```{r}
sdrf_has_repeated_columns(doc)
head(sdrf_records(doc)[, c("source name", "characteristics[organism]")])
```

A ragged document, from mzLib's own fixtures:

```{r}
ragged <- sdrf_read("PXD059974.sdrf.tsv")
sdrf_ragged_row_count(ragged)
```

## Pool several into one analysis table

Name the documents: the names become each row's provenance label.

```{r}
pooled <- sdrf_pool(c(malaria = "PXD000070.sdrf.tsv", colon = "PXD026824.sdrf.tsv"), limit = 4)
pooled
sdrf_source_documents(pooled)
```

Columns are the union over the documents, and a cell a document did not have is filled with
`"not available"` - so a fill rate computed on the pooled table is not the fill rate of the
originals. The pooled table is for analysis, not for deposit: two experiments may both have a
"Sample 1", so keep `sdrf_source_documents()` in any key.

## Instead of `readers_read_records()`

`readers_read_records()` recognises SDRF too, but joins each row into one semicolon-separated
string, and SDRF's own `NT=...;AC=...` grammar puts semicolons inside cells. Use `sdrf_read()`.

```{r teardown, include = FALSE}
mzLibR:::replay_bridge_stop(.mzlibr_example)
```
